Foreground Signatures and Mitigation in Neutral Hydrogen Power Spectra From The Reionization Epoch

### 2014-10-17 foreground_delay_spectrum_modeling.tex
From 1d1c070: ============
(*) Made minor changes based on Abraham's comments.

Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components most likely to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a larger role than previously thought. % We find that the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon. We find that diffuse Galactic emission plays a larger role then previously thought. 
We quantify this contamination by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The MWA passband of width 30.72~MHz is divided coarsely into 24$\times$1.28~MHz sub--bands with each sub--band weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. After correcting for the shape of these coarse channels, the fine channels (40~kHz width) at the edges of these sub--bands are flagged because they are known to be possibly contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

@@

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=95$~K per polarization for all frequency channels for all antenna pairs throughout the course of the observation. % We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction. ({\bf Since the bandpass correction is digital, shouldn't the noise stay flat after bandpass correction? -- JDB: yes})

@@

The observations shown here were undertaken on Aug 23, 2013. This day has been chosen as a reference data set for comparison within the MWA EoR collaboration and was selected on the basis of a low number of known system errors from among a long stretch of stable observing.

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by the additional weight of the flagged channels which occur rarely for interference and regularly every 1.28~MHz where the edges of coarse channels are known to have a slight aliasing effect. In delay--space, these weights translate into a convolution by a grated point spread function (PSF) with gratings at multiples of $\eta=0.78\,\mu$s. We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

The MWA has a very wide field of view ($\gtrsim 20$\arcdeg~ at 185~MHz) and has non--zero response all the way to the horizon. As already noted, foreground objects fill the {\it foreground wedge} with an amplitude which depends on the tile power gain in that direction, with objects at the horizon being closest to the boundaries of the {\it wedge}.

@@

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. Since this map is predominantly based on the 408~MHz all--sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. We model the diffuse foreground spectra with a single spectral index at each pixel in the map, found by comparing model maps at 170~MHz and 200~MHz.

@@

It may be noted that numerous features overlap at varying levels of significance depending on combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna spacing length and orientation, power--pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectra of foreground emission.

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\overline{\mathbf{x}}|\le$~200~m). This is contrary to expectations because longer baselines are not expected to be sensitive to diffuse emission. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.   %% This contributes to a {\it pitchfork}--shaped feature which will be discussed later in \S\ref{sec:composite} in more detail.

@@

When not dominated by the bright emission from the Galactic center, which happens only for certain pointings at certain times, the delay spectrum of the combined foreground model is dominated, in somewhat equal measure, by diffuse and compact emission. This is illustrated by a more detailed examination of the {\it zenith} pointing in our study.

@@

In our example, we discard eastward antenna spacings (-22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5) of lengths $|\overline{\mathbf{x}}| < 40$~m. Figure~\ref{fig:before-after} shows the delay spectra obtained with all antenna spacings (top panel) and after applying our screening technique (bottom panel) on the {\it off--zenith} observation. Notice the remarkable reduction --- by about an order of magnitude --- in foreground contamination in this region after application of our screening technique.

@@

This screening technique can be customized when it is necessary to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\overline{\mathbf{x}}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\overline{\mathbf{x}}|>40$~m has no effect on further foreground contamination and would only lead to loss of sensitivity. On the other hand, a bright compact object at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or all of the eastward oriented antenna spacings.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/baseline_screening_nside_64_Tsys_95.0K.eps}
\caption{Foreground contamination in {\it EoR window} for the {\it off--zenith} pointing as a function of orientation and length of discarded antenna spacings. Antenna spacings oriented eastward with varying degrees of directedness are considered: -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy the above criteria, ones with lengths $\le |\overline{\mathbf{x}}|_\textrm{max}$ ($x$--axis) are discarded. Lower (dashed line) and upper (dot--dashed) limits on foreground contamination are derived from northward (67\fdg 5~$\le\theta_\textrm{b}<$~112\fdg 5) and eastward ($|\theta_\textrm{b}|\le$~22\fdg 5) antenna spacings respectively. Foreground contamination in the {\it EoR window} reduces by factor $\sim$1.5 ($|\theta_\textrm{b}|\le$~7\fdg 5) to $\sim$10 ($|\theta_\textrm{b}|\le$~22\fdg 5). Short baselines have maximum effect on foreground contamination. For instance, discarding antenna spacings oriented $|\theta_\textrm{b}|\le$~22\fdg 5 with lengths $|\overline{\mathbf{x}}|>40$~m has no effect in further reducing foreground contamination. \label{fig:screening}}
\end{figure}

>>>>

Considerable effort has been made towards understanding the $k$--space behavior of foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,liu11,dat10,bow09,liu09}. A general model of scattering of smooth spectrum foregrounds to higher $k$--modes by the wide-field response of the instrument, into the so-called `wedge', has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to discard $k$--modes which could be contaminated ({\it avoidance}). The more aggressive alternative is to subtract a sky model and regain access to modes lost by {\it avoidance}. In both cases, which parts of the sky are most critical to either {\it avoid} or {\it subtract} has remained largely uncertain. Here, we will focus primarily on extending the {\it avoidance} strategy by identifying foreground components at greatest risk to ``leak'' from foreground modes to EoR modes and demonstrating a scheme for down--weighting these components.

@@

Matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds that most severely obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a larger role than previously thought. % We find that the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon. We find that diffuse Galactic emission plays a larger role then previously thought. 
We quantify this contamination by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/fourier_space_185.0_MHz_30.7_MHz.eps}
\caption{Illustration of Fourier space in which delay (and power) spectra of H{\sc i} signals from the EoR are presented. The $x$--axis is denoted by $|\overline{\mathbf{x}}|$ (antenna spacing) or $k_\perp$ (transverse wavenumber). The $y$--axis denoted by $\eta$ (delay) or $k_\parallel$ (line--of--sight wavenumber). The dark shaded region is referred to as the {\it foreground wedge} where smooth spectrum foregrounds reside. Its boundaries (solid lines), given by light travel time for corresponding antenna spacings, are referred to as horizon delay limits. Regions outside the {\it wedge} and the narrow white unshaded strips (caused by convolution with the instrumental transfer function) are expected to be relatively free of foreground contamination. Due to an undesirable grating pattern (dot--dashed lines) of the instrument used in this study, namely the MWA, we conservatively identify a restricted region of high EoR sensitivity (medium shade) and refer to it as the {\it EoR window} in this paper. \label{fig:fourier-space}}
\end{figure}

@@

The MWA construction was completed in 2012 and, after commissioning, began its EoR observing program in 2013. In its final configuration the MWA is a 128--tile interferometer array capable of observing a 30.72~MHz instantaneous band anywhere in the range 80--300~MHz. Each tile is a phased array of 16 dipoles, each in the shape of a bow--tie. This yields a primary field of view $\gtrsim$~20\arcdeg~ wide and multiple secondary lobes. The array is arranged as a centrally condensed core of $\sim$~300~m --- there are many spacings in the range 5--50~m --- and a radial density that falls off as the inverse of the radius, with the longest baselines at 3~km \citep{bea12}. Here we focus on antenna spacings $|\overline{\mathbf{x}}| \le 200$~m (spatial scales relevant to reionization) and a frequency range 170--200~MHz. The center frequency of 185~MHz corresponds to a redshift, $z\simeq 6.68$.

The MWA passband of width 30.72~MHz is divided coarsely into 24$\times$1.28~MHz sub--bands with each sub--band weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. After correcting for the shape of these coarse channels, the fine channels (40~kHz width) at the edges of these sub--bands are flagged because they are known to be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

@@

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=95$~K per polarization for all frequency channels for all antenna pairs throughout the course of the observation. Estimation of $T_\textrm{sys}$ from MWA data is a subject of active investigation. In our work, we have chosen a $T_\textrm{sys}$ that matches the thermal noise observed in the data. % We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction. ({\bf Since the bandpass correction is digital, shouldn't the noise stay flat after bandpass correction? -- JDB: yes})

@@

The observations shown here were undertaken on Aug~23, 2013. Data set from this day has been chosen as a reference for comparison within the MWA EoR collaboration because this is one of the days featuring a low number of known system errors from among a long stretch of stable observing.

@@

We obtain the delay spectrum of these calibrated visibilities taking the delay transform of each baseline's spectrum (equation~\ref{eqn:delay-transform}) choosing $W'_f(f)$ to be a {\it Blackman--Harris} window function. The spectrum is multiplied by the additional weight of the flagged channels which occur rarely for interference and regularly every 1.28~MHz where the edges of coarse channels are known to have a small aliasing effect. In delay--space, these weights translate into a convolution by a grated point spread function (PSF) with gratings at multiples of $\eta=0.78\,\mu$s. We deconvolve this PSF using a one dimensional CLEAN algorithm \citep{tay99} as described for the delay axis by \citet{par09,par12}. The CLEAN procedure iteratively finds and subtracts peak values convolved by the Fourier transform of the weights. We limit the selection of peaks to modes inside the horizon delay limit, corresponding to smooth spectrum objects in the visible sky hemisphere.

@@

The MWA has a very wide field of view ($\gtrsim 20$\arcdeg~ at 185~MHz) and has non--negligible response all the way to the horizon. As already noted, foreground objects fill the {\it foreground wedge} with an amplitude which depends on the tile power gain in that direction, with objects at the horizon being closest to the boundaries of the {\it wedge}.

@@

For the diffuse component, we use an all--sky radio foreground model \citep{deo08} to estimate the emission at 185~MHz. At this frequency, since this map is predominantly based on the 408~MHz all--sky map of \citet{has82} which has an angular resolution of 0\fdg 85, we smoothed the 185~MHz map to the same resolution. However, to avoid any artifacts from sampling this map, we sample it at $\approx 27$\arcmin~intervals. We model the diffuse foreground spectra with a single spectral index at each pixel in the map, found by comparing model maps at 170~MHz and 200~MHz.

@@

Note that numerous features may overlap at different degrees of significance depending on combinations of parameters. We assign the features to their predominant causes. Secondly, we have used noiseless cases to clearly illustrate the observed foreground signatures. With the addition of noise in the visibilities which is subject to observing time, some of the weaker features may not be as prominently visible. Since the foreground signatures are far too numerous and they reside in a vast volume of parameter space such as antenna spacing length and orientation, power--pattern, patch of sky under observation, and instrumental configuration, we highlight only some examples of the most notable features observed in the delay spectra of foreground emission.

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\overline{\mathbf{x}}|\le$~200~m). This is contrary to expectations because these longer baselines are not expected to be sensitive to diffuse emission (on scales $\gtrsim$~0\fdg 85) of our input foreground model. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.   %% This contributes to a {\it pitchfork}--shaped feature which will be discussed later in \S\ref{sec:composite} in more detail.

@@

When not dominated by the bright emission from the Galactic center, which happens only for certain pointings at certain times, the delay spectrum of the combined foreground model is composed of diffuse and compact emission, in roughly equal measure. This is illustrated by a more detailed examination of the {\it zenith} pointing in our study.

@@

In our example, we discard eastward antenna spacings (-22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5) of lengths $|\overline{\mathbf{x}}| < 40$~m. Figure~\ref{fig:before-after} shows the delay spectra obtained with all antenna spacings (top panel) and after applying our screening technique (bottom panel) on the {\it off--zenith} observation. Notice the remarkable reduction --- by about an order of magnitude --- in foreground contamination in the {\it EoR window} after application of our screening technique. The color scale range has been adjusted relative to the previous figures to show this significant reduction in foreground contamination.

@@

This screening technique can be generalized to optimize between foreground mitigation and loss of sensitivity from discarding data. Figure~\ref{fig:screening} shows how the typical foreground contamination in the {\it EoR window} depends on the orientations and lengths of discarded antenna spacings. We choose antenna spacings oriented eastward to varying degrees of directedness, i.e., -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy these criteria, we discard data from those whose lengths are shorter than $|\overline{\mathbf{x}}|_\textrm{max}$ ($x$--axis) and show foreground contamination estimated in the {\it EoR window} from all remaining antenna spacings. In other words, these plots demonstrate the progress in mitigation as orientation and maximum length of discarded antenna spacings are varied. Foreground contamination estimated from only northward (dashed line) and eastward (dot--dashed line) antenna spacings represent the lower and upper limits, respectively, and are shown for reference. It is seen that foreground contamination can be mitigated by a factor between $\sim 1.5$ ($|\theta_\textrm{b}|\le$~7\fdg 5) and $\sim 10$ ($|\theta_\textrm{b}|\le$~22\fdg 5). It is notable that short antenna spacings play maximum role in determining foreground contamination. For instance, among antenna spacings with orientations $|\theta_\textrm{b}|\le$~22\fdg 5, discarding those with lengths $|\overline{\mathbf{x}}|\gtrsim 30$~m does not mitigate foreground contamination any further and would only lead to loss of sensitivity. On the other hand, a bright compact object at the same location will give rise to foreground contamination even on longer antenna spacings. Such cases will necessitate discarding more or all of the eastward oriented antenna spacings.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/baseline_screening_nside_64_Tsys_95.0K.eps}
\caption{Foreground contamination in {\it EoR window} for the {\it off--zenith} pointing as a function of orientation and length of discarded antenna spacings. Antenna spacings oriented eastward with varying degrees of directedness are considered: -7\fdg 5~$\le\theta_\textrm{b}<$~7\fdg 5 (solid circles), -15\arcdeg~$\le\theta_\textrm{b}<$~15\arcdeg~ (solid squares), and -22\fdg 5~$\le\theta_\textrm{b}<$~22\fdg 5 (solid stars). Among antenna spacings that satisfy the above criteria, ones with lengths $\le |\overline{\mathbf{x}}|_\textrm{max}$ ($x$--axis) are discarded. Lower (dashed line) and upper (dot--dashed) limits on foreground contamination are derived from northward (67\fdg 5~$\le\theta_\textrm{b}<$~112\fdg 5) and eastward ($|\theta_\textrm{b}|\le$~22\fdg 5) antenna spacings respectively. Foreground contamination in the {\it EoR window} reduces by factor $\sim$1.5 ($|\theta_\textrm{b}|\le$~7\fdg 5) to $\sim$10 ($|\theta_\textrm{b}|\le$~22\fdg 5). Short baselines have maximum effect on foreground contamination. For instance, discarding antenna spacings oriented $|\theta_\textrm{b}|\le$~22\fdg 5 with lengths $|\overline{\mathbf{x}}|\gtrsim 30$~m has no effect in further reducing foreground contamination. \label{fig:screening}}
\end{figure}
